Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-20 Direction_v2.tex
The updates from Henry. --Lindley

\documentclass[aps,prc,twocolumn,groupedaddress,showpacs,amsmath,amssymb,floatfix,superscriptaddress]{revtex4}
\usepackage{multirow}
\usepackage{graphicx,subfigure}
\usepackage{times}
\usepackage{textcomp}
\usepackage{hyperref}

@@

\newcommand{\nuebar}{$\overline{\nu}_{e}$}
\newcommand{\PerTonDay}{(ton$\cdot$day)$^{-1}$}
\hyphenation{KamLAND}

@@

\begin{abstract}
Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would greatly enhance the scientific reach of these detectors,
especially for searches for neutrinoless double-beta decay. In this
paper, we develop a technique for extracting particle direction and
evaluate different detector advances that could be used to make
direction reconstruction a reality in a kiloton-scale detector.
\end{abstract}

@@

\section{Introduction}
Liquid scintillator based detectors are responsible for several of the
critical measurements that have determined our present understanding
of neutrino masses and mixings. These measurements include KamLAND's
measurement of reactor antineutrino oscillation at a distance of
$\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar
neutrino oscillation\cite{borexino}, and most recently the short
baseline reactor antineutrino experiments that measured oscillations
due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo},
Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.
Scintillator-based neutrino detectors will continue to be important for the
next set of neutrino measurements, from the determination of the
neutrino mass hierarchy\cite{juno,reno50} to elastic scattering
measurements\cite{isodarscatt} and sterile neutrino
searches\cite{isodar,nist} and for non-proliferation
applications\cite{nucifer, songs}.

The scalability of these detectors to large volumes also makes them
highly competitive for neutrinoless double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a
positron-electron pair with energies in the few MeV range.  Currently
one of the best limits for the $0\nu\beta\beta$ half-life comes from
KamLAND-Zen\cite{KZ0nu}.

@@

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly the the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiationless transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The molecules in liquid scintillators are not isolated. Radiationless
processes transfer energy between the molecules. The probability of
energy transfer between molecules increases as the overlap between the
molecule's absorption and emission spectra increase. The absorption
and emission spectra overlap at some level for all
molecules. Consequently if there is only one type of molecule in the
scintillator cocktail, the light output would be reduced due to
inefficiencies in the energy transfer through multiple absorption and
reemission processes. Aromatic solutes or fluors are added to the
primary solvent to shift the wavelength of the photons to longer
wavelengths where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of
PPO as the fluor. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone pointed in the direction of the particles trajectory and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
In Geant4 optical photons are assigned the group velocity in the wavelength 
region of normal dispersion which is relevant to our study. 
The longer wavelengths Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelength, it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light in time and then reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directionality reconstruction in liquid scintillators, a Geant4 \cite{geant4one,geant4two} simulation has been set up. The presented simulation has been developed using standard Geant4 (version 4.9.6) functionality for the liquid scintillator optical model. The simulation has been built from a customized official Geant4 application example and it is kept rather generic to provide generally applicable information about the main factors in directionality reconstruction.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each (equally divided energy of $^116$Cd $0\nu\beta\beta$ decay) at the center of the sphere with initial directions along the x and -x-axis. Only Cerenkov photons are drawn to illustrate the directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The implemented detector geometry is a sphere of 6.5~m diameter filled with liquid scintillator (see Figure \ref{detector_view}). Default scintillator properties have been chosen to match the KamLAND scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene (1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The implemented scintillator properties include the atomic composition and density ($\rho$ = 0.78~g/ml), the wavelength-dependent attenuation length \cite{tajimaMaster} and refractive index \cite{OlegThesis}, the scintillation emission spectrum, emission rise time ($\tau_r$ = 1.0~ns) and emission decay time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 0.13) \cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV) and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV) \cite{ChrisThesis}. Variations from the baseline KamLAND case are discussed below when applicable. Reflections, reemission of absorbed photons in the scintillator bulk volume and scattering has not been included. These effects are expected to be corrections to the main effects studied in this paper and are subject of future work.

The inner sphere surface is used as the photodetector. It is treated as fully absorbing (no reflections) with a photodetector coverage of 100~\%. Two important photodetector properties have been varied: Transit time spread (TTS, default $\sigma$ = 0.1~ns) and wavelength-dependent quantum efficiency (QE) for photoelectron production. Default is the QE of a bialkali photocathode (Hamamatsu R7081 PMT), digitized values come from the Double Chooz \cite{dctwo} Monte Carlo simulation. We note that the KamLAND 17 inch PMTs use the same photocathode type (bialkali) with similar quantum efficiency.

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with different settings. PEs from Cerenkov light (black) and scintillation light (grey) are compared. The dashed vertical line illustrates a time cut at 34~ns. (a) Default simulation: bialkali photocathode and TTS = 0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs from scintillation and 108~PEs from Cerenkov light. (b) Default simulation settings except for TTS = 1.28~ns (KamLAND 17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from scintillation and 88~PEs from Cerenkov light. (c) Default simulation settings except for a GaAsP photocathode. After the 34.0~ns time cut we get 226~PEs from scintillation and 229~PEs from Cerenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}
Before the simulation results for different simulation settings are discussed in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation light emission. Past neutrino experiments were not highly interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. We assume a rise time of 1.0~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector at the level of precision needed for direction reconstruction. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (see Equation (\ref{eqGroup})) increases with increasing photon wavelengths for normal dispersion. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and applied QE but no TTS. The true hit time distributions of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are on average created about 0.5~ns earlier than PEs from scintillation light. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time.

On the detector side, the time information of single photoelectrons is affected by the TTS of the photodetectors, a number which can be different by orders of magnitude depending on the detector type. The default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with the large area picosecond photodetectors (LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors (HPDs)\cite{hpdThesis}. In fact, even significantly lower TTS numbers are realistic with the LAPPD. Finally, the accuracy of the vertex reconstruction is reflected in broadening and distortions of the time spectrum (after time-of-flight correction has been applied). Another effect which is to be addressed in future work is the granularity of the photodetectors which adds to the uncertainty in the light path and thus the time-of-flight correction.

In Sections \ref{detector_timing_sec} to \ref{scintillator_emission_sec}, we comparatively study the photoelectron timing for different detector configurations. We follow the idea to increase discrimination between Cerenkov and scintillation light by accurate detector timing. The main relevant quantities provided by the Geant4 simulation are the photoelectron hit positions and the detection times after TTS has been applied. In section \ref{reconstruction_sec} these quantities are used for event reconstruction.

@@

First we discuss results for the default simulation settings which are described in the previous section. Figure \ref{time_plots_comparison} (a) shows the TTS-smeared photoelectron detection times for 1000 simulated electrons with 5~MeV energy in the center of the detector with initial momentum direction coinciding with the x-axis. The photoelectrons induced by Cerenkov light arrive earlier, as expected due to the instantaneous emission and the higher average photon speed (higher wavelengths) compared to scintillation after QE weighting. There is however significant overlap with the PE time distribution coming from scintillation light. In order to compare simulations with different parameters to each other a fixed time cut $t \leq$ 34.0~ns (or 33.0~ns) is applied using the truth information of the electron starting time instead of an event-by-event time reconstruction (see Section \ref{reconstruction_sec} for reconstruction results). For the default simulation case, the number of PEs per event coming from Cerenkov light after the 34.0~ns time cut (108) is 98~\% of the total number of PEs from Cerenkov light (110). For scintillation the number of PEs after the time cut (171) is only 3.2~\% of the total scintillation-induced PEs (5445). This indicates the effectiveness of a time cut to separate Cerenkov light from scintillation light. An important figure of merit when comparing different simulation settings is the ratio of Cerenkov/scintillation induced PE after the 34.0~ns time cut since a higher ratio means more directional information per PE. The ratio is 0.63 for the default settings. Note that this ratio depends on the time cut, which will be optimized in future work together with the absolute number of PEs per event passing the cut. 
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{Angular distribution of photoelectron hits relative to the original electron direction along the x-axis: $\cos(\Psi) = x_{hit}/\vec{r}_{hit}$. 1000 electron events with 5~MeV at the center are summed. Default simulation settings are used and both Cerenkov and scintillation light are included. A 33.0~ns time cut (black) is compared with a 34.0~ns time cut (grey). \label{cerenkov_cone}}
        \end{center}
\end{figure}

Figure \ref{cerenkov_cone} displays the PE hit pattern after cutting on time. Although this time cut is a simplification of actual time reconstruction effects, we can use it to indicate the spatial distribution of hits after timing information has been used to separate Cerenkov and scintillation light. The Cerenkov ring structure can be clearly seen (peak at about 46\textdegree) which demonstrates that the directional signal conveyed by the Cerenkov photons is not erased by scattering of the initial 5~MeV electrons.

If the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS = 1.28~ns) are used in the simulation, the broadening of the time distributions leads to a strongly decreased ratio of Cerenkov over scintillation light (0.25) after the 34~ns time cut (see FIG. \ref{time_plots_comparison} (b)). This shows that low photodetector TTS is critical for directionality reconstruction and motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
Complementary to lowering the photodetector TTS, optimization of the wavelength-dependent QE is promising. Since Cerenkov photons which passed 6.5~m of scintillator have higher average wavelengths than scintillation photons, a photodetector which is more sensitive at high wavelengths increases not only the absolute number of PEs but also the ratio between Cerenkov- and scintillation-induced PEs. Therefore, a simulation has been run with the QE of an extended red-sensitive GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).

Figure \ref{time_plots_comparison} (c) shows the results for the modified simulation with high QE in the red spectral region. The higher absolute number of PE coming from Cerenkov light (factor of $\approx$ 2) and the increased Cerenkov/scintillation ratio (1.6) after the time cut would significantly enhance the directionality reconstruction capabilities given that high coverage can be achieved with red-sensitive photocathodes.

@@

In order to isolate the effect of a different emission spectrum, the other simulation settings, including the KamLAND absorption spectrum are kept unchanged. Compared to the default case shown in FIG. \ref{time_plots_comparison} (a) the separation is worse (as expected) because the scintillation light wavelengths are higher than in the KamLAND emission spectrum. However, advances in the production of commercial quantum dot samples could yield quantum dots which have similar, single peak emission shapes at lower wavelengths. This case has been simulated using the same spectral shape of the measured core-shell quantum dot emission but shifted to lower wavelengths such that the emission peak is centered at 384~nm. This peak emission value has been measured for other types of QDs, however with a much more pronounced tail \cite{qdot2}. The resulting PE time distribution shows improved separation of Cerenkov and scintillation light compared to the default simulation. After the 34.0~ns cut on the TTS-smeared PE time we obtain a higher Cerenkov/scintillation ratio of 0.86 (107 PE from Cerenkov light and 124 PE from scintillation). The number of Cerenkov-induced PE after the time cut is unchanged while the number of PEs coming from scintillation light is decreased due to the higher average photon travel times.

@@

The first step of the reconstruction process relies on exact numerical calculations of vertex candidates from quadruplets of hits. Given a single point source, we need four constraints to solve for the four unknowns of the vertex ~\cite{Smy}. This approach would provide an exact solution in the case of four prompt, unscattered photons originating from a common point. However, many of these randomly chosen quadruplets will produce anomalous solutions due to real world effects such as delayed emission and deviations from the point-like geometry. Nonetheless, we found that any chosen subset of 400 quadruplets was a sufficiently large ensemble to assure that some solutions will be close to the true vertex.

>>>>

\documentclass[aps,prc,twocolumn,groupedaddress,showpacs,amsmath,amssymb,floatfix,superscriptaddress]{revtex4}
\usepackage{multirow}
\usepackage{graphicx,subfigure}
\usepackage{times}
\usepackage{textcomp}
\usepackage{hyperref}
\usepackage{lineno}

@@

\newcommand{\nuebar}{$\overline{\nu}_{e}$}
\newcommand{\PerTonDay}{(ton$\cdot$day)$^{-1}$}
\newcommand{\GEANT}{GEANT4}
\hyphenation{KamLAND}

\linenumbers

@@

\begin{abstract}
Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would enhance the scientific reach of these detectors,
especially for searches for neutrino-less double-beta decay. In this
paper, we develop a technique for extracting particle direction and
evaluate different detector advances that could be used to make
direction reconstruction a reality in a kiloton-scale detector.
\end{abstract}

@@

\section{Introduction}
Liquid scintillator based detectors are responsible for several of the
critical measurements that have determined our present understanding
of neutrino masses and mixings. These measurements include KamLAND's
measurement of reactor anti-neutrino oscillation at a distance of
$\sim$200~km\cite{kam2013}, Borexino's measurement of $^{7}$Be solar
neutrino oscillation\cite{borexino}, and most recently the short
baseline reactor anti-neutrino experiments that measured oscillations
due to $\theta_{13}$ at a distance of 1~km: Daya Bay\cite{dbtwo},
Double Chooz\cite{dctwo, dchydrogen}, and RENO\cite{reno}.
Scintillator-based neutrino detectors will continue to be important for the
next set of neutrino measurements, from the determination of the
neutrino mass hierarchy\cite{juno,reno50} to elastic scattering
measurements\cite{isodarscatt} and sterile neutrino
searches\cite{isodar,nist}, and for non-proliferation
applications\cite{nucifer, songs}.

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  Currently one of the
best limits for the $0\nu\beta\beta$ half-life comes from
KamLAND-Zen\cite{KZ0nu}.

@@

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiation-less transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The molecules in liquid scintillators are not isolated. Radiation-less
processes transfer energy between the molecules. The probability of
energy transfer between molecules increases as the overlap between the
molecule's absorption and emission spectra increase. The absorption
and emission spectra overlap at some level for all
molecules. Consequently if there is only one type of molecule in the
scintillator cocktail, the light output would be reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluors are added to the
primary solvent to shift the wavelength of the photons to longer
wavelengths where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of
PPO as the fluor. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone centered on the direction of the particle trajectory, and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by~\cite{Cherenkov_spectrum}:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directional
reconstruction in liquid scintillators, a 
\GEANT~\cite{geant4one,geant4two} simulation has been set up. The 
simulation has been developed using \GEANT Version 4.9.6 for the
liquid scintillator optical model, in which optical photons are
assigned the group velocity in the wavelength region of normal
dispersion.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each
        (equally divided energy of $^{116}$Cd $0\nu\beta\beta$ decay) at
        the center of the sphere with initial directions along the x
        and -x-axis. Only Cerenkov photons are drawn to illustrate the
        directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The detector geometry is a sphere of 6.5~m diameter filled
with liquid scintillator doped with $^{116}$Cd (see Figure
\ref{detector_view}). The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100~\%. Two important photodetector properties have been varied: 1)
the transit time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT), for which digitized values come from the Double Chooz \cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default
        simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS
        (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive
        photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{Photo-electron (PE) arrival times after application
        of the transit-time spread (TTS) for the simulation
        of 1000 electrons (5~MeV) with different values of the TTS and
        wavelength response. PEs from
        Cerenkov light (black, solid line) and scintillation light
        (red, dotted line) are
        compared. The dashed vertical line illustrates a time cut at
        34~ns. (a) Default simulation: bialkali photocathode and TTS =
        0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs
        from scintillation and 108~PEs from Cerenkov light. (b)
        Default simulation settings except for TTS = 1.28~ns (KamLAND
        17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from
        scintillation and 88~PEs from Cerenkov light. (c) Default
        simulation settings except for a GaAsP photocathode. After the
        34.0~ns time cut we get 226~PEs from scintillation and 229~PEs
        from Cerenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}

Before the simulation results for different simulation settings are
discussed in the following sections, we highlight the effects which
contribute to the timing of the scintillator detector system. First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns, more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5~m diameter detector at the level of precision needed
for direction reconstruction. Due to the wavelength-dependence of the
refractive index the speed of light in the scintillator (see Equation
(\ref{eqGroup})) increases with increasing photon wavelengths for
normal dispersion. In order to study the size of this effect, we
extracted results from a simplified simulation of 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including
the transit-time spread. The true hit time distributions of
photoelectrons were analyzed for scintillation light and Cerenkov
light separately. Photoelectrons coming from Cerenkov light are on
average created about 0.5~ns earlier than PEs from scintillation
light. The RMS values from PE time distributions for Cerenkov and
scintillation light are both about 0.5~ns. Note that these numbers
include the effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the TTS of the photodetectors, a number which can be
different by orders of magnitude depending on the detector type. The
default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with
the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD. The uncertainties in the vertex
reconstruction are reflected in broadening and distortions of the time
spectra.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We follow
the idea to increase discrimination between Cerenkov and scintillation
light by accurate detector timing. The main relevant quantities
provided by the \GEANT simulation are the photoelectron hit positions
and the detection times after the TTS resolution has been applied. In
Section~\ref{reconstruction_sec} these quantities are used for event
reconstruction.

@@

We first discuss results for the default simulation settings described
in the previous section. Figure \ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cerenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed longer
wavelengths) compared to scintillation light. There is however
significant overlap of the two arrival time distributions.

In order
to compare simulations with different parameters, a fixed time cut $t
\leq$ 34.0~ns  is applied using the truth information 
instead of an event-by-event time reconstruction (see Section
\ref{reconstruction_sec} for reconstruction results). For the default
simulation case, the number of PEs per event coming from Cerenkov
light after the 34.0~ns time cut (108) is 98~\% of the total number of
PEs from Cerenkov light (110). For scintillation the number of PEs
after the time cut (171) is only 3.2~\% of the total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cerenkov light from scintillation light. An
important figure of merit when comparing different simulation settings
is the ratio of Cerenkov/scintillation induced PE after the 34.0~ns
time cut since a higher ratio means more directional information per
PE. The ratio is 0.63 for the default settings. Note that this ratio
depends on the time cut, which will be optimized in future work
together with the absolute number of PEs per event passing the cut.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cerenkov and
        scintillation light are included. } 
        \label{cerenkov_cone}
        \end{center}
\end{figure}

Figure \ref{cerenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits after timing information has been used to
separate Cerenkov and scintillation light. The Cerenkov ring structure
can be clearly seen in the peak at about 46\textdegree, demonstrating
that the directional signal conveyed by the Cerenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cerenkov over
scintillation light (0.25) after the 34~ns time cut (see
FIG. \ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to improving the photodetector TTS, optimization of the
wavelength-dependent QE is promising. Since Cerenkov photons which
passed 6.5~m of scintillator have higher average wavelengths than
scintillation photons, a photodetector which is more sensitive at high
wavelengths increases not only the absolute number of PEs but also the
ratio between Cerenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).  Figure
\ref{time_plots_comparison} (c) shows the results for the modified
simulation with high QE in the red spectral region. The higher
absolute number of photo-electrons coming from Cerenkov light (factor
of $\approx$ 2) and the increased Cerenkov/scintillation ratio (1.6)
after the time cut would significantly enhance the directionality
reconstruction.

@@

In order to isolate the effect of a different emission spectrum, the
other simulation settings, including the KamLAND absorption spectrum,
are kept unchanged. Compared to the default case shown in
FIG. \ref{time_plots_comparison} (a) the separation is worse (as
expected) because the scintillation light wavelengths are higher than
in the KamLAND emission spectrum. However, advances in the production
of commercial quantum dot samples could yield quantum dots which have
similar, single peak emission shapes at lower wavelengths. This case
has been simulated using the same spectral shape of the measured
core-shell quantum dot emission but shifted to lower wavelengths such
that the emission peak is centered at 384~nm. This peak emission value
has been measured for other types of QDs, however with a much more
pronounced tail \cite{qdot2}. The resulting PE time distribution shows
improved separation of Cerenkov and scintillation light compared to
the default simulation. After the 34.0~ns cut on the TTS-smeared PE
time we obtain a higher Cerenkov/scintillation ratio of 0.86 (107 PE
from Cerenkov light and 124 PE from scintillation). The number of
Cerenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

@@

The first step of the reconstruction process relies on exact numerical
calculations of vertex candidates from quadruplets of hits. Given a
single point source, we need four constraints to solve for the four
unknowns of the vertex ($x,y,z, and t_0$)~\cite{Smy}. This approach
would provide an exact solution in the case of four prompt,
unscattered photons originating from a common point. However, many of
these randomly chosen quadruplets will produce anomalous solutions due
to real world effects such as delayed emission and deviations from the
point-like geometry. Nonetheless, we found that any chosen subset of
400 quadruplets was a sufficiently large ensemble to assure that some
solutions will be close to the true vertex.
